% ── ABSTRACT, Foundations of Physics variant (~250-word target, trimmed from
%    the canonical ~300-word abstract in 00-abstract.tex) ─────────────────

Three independent traditions --- quantum physics, Whitehead's process
philosophy, and the \bahai writings of \Bahaullah and \AbdulBaha --- converge
on the same structural description of reality: existence is constituted by
relationship, properties emerge only through relational interaction, and the
force sustaining those relationships is, in each tradition's vocabulary,
love.

This paper makes three arguments.
First, local substance ontology is \emph{formally incompatible} with quantum
mechanics: Bell's theorem \citep{Bell1964} and the experimental violation of
its inequalities \citep{Aspect1982,Hensen2015} undermine the assumption that
systems carry intrinsic properties prior to measurement, making Whitehead's
process ontology a systematically developed alternative framework for
quantum reality.
Second, the \bahai writings constitute an \emph{independent prior
instantiation} of process ontology, articulating its essential claims
decades before \citet{Whitehead1929} and roughly a century before physics
confirmed it experimentally.
Third, these findings establish a \emph{triadic convergence}: three
traditions, three methodologies, across 170 years, detecting the same
feature of reality.

The convergence is formalized through six structural isomorphisms
(relational ontology, nonlocal interconnection, observer-dependence,
decoherence, holographic order, and motion as life) and the Relational
Coherence Framework (\RCF), developed in a companion physics
paper~\citep{Adams2026Physics}, which translates the philosophical claim
into testable quantum-mechanical predictions. Each tradition validates its
own thesis independently; the convergence is an independent confirmation by
separate instruments of the same underlying structure. In each tradition's
vocabulary the relational principle sustaining this structure is named
\emph{love} --- \mahabbat in the \bahai lexicon, the creative advance in
Whitehead, and the nonlocal coherence revealed by entanglement --- and this
convergence on a single substantive identification is itself the central
finding.
