% ── SECTION 6: THE SEMANTIC BRIDGE (~600 words) ──────────────────────────────

\section{The Semantic Bridge: One Phenomenon, Many Names}
\label{sec:semantic}

The central interpretive claim of this paper is this:
\mahabbat is not a human emotion projected onto an indifferent universe.
It is the fundamental property of reality --- the force that forms and sustains
relationship at every scale --- which human beings experience, in its most
conscious and intense form, as emotion.

Table~\ref{tab:semantic} presents the semantic bridge:
each tradition's technical term for the same phenomenon, with the key source
and the thesis it supports.

\begin{table}[h]
\centering
\caption{The semantic bridge: one phenomenon, many names.}
\label{tab:semantic}
\begin{tabularx}{\textwidth}{lXXl}
\toprule
\textbf{Domain} & \textbf{Technical Term} & \textbf{What It Names} & \textbf{Thesis} \\
\midrule
Quantum mechanics   & Entanglement / Affinity Tensor $\alpha$
                    & Nonlocal relational bond; measure of coherence & T1 \\
Process Philosophy  & Prehension / Eros
                    & Relational act of becoming; creative drive toward richer relation & T2 \\
\bahai Writings     & \mahabbat / Affinity
                    & Creative foundation of all existence at every degree & T3 \\
Human experience    & Love
                    & Most conscious form of the universal relational force & All \\
\bottomrule
\end{tabularx}
\end{table}

Note what is \emph{not} being claimed.
We are not asking physicists to call entanglement ``love,'' or \bahai scholars
to redefine \mahabbat as the Affinity Tensor.
Each tradition keeps its own vocabulary, its own methodology, and its own
epistemic standards.
What we claim is that when each tradition's instruments point at the same
structure, from independent directions, across a 170-year sequence,
that structure is real.

More precisely: the convergence is functional, not nominal.
What entanglement \emph{does} in quantum mechanics --- maintain a single
quantum state across spatial separation, preserve identity through
relational coupling, sustain non-local coherence against environmental
decoherence --- is the same work that \mahabbat \emph{does} in the
\bahai writings: sustain unified being through the affinity and agreement
that hold parts in coherent integration, such that the failure of
affinity is identified by \AbdulBaha as the very meaning of nonexistence.
Whiteheadian prehension performs the same work in process metaphysics:
carry the unity of prior occasions forward into the present event of
becoming, integrating a plurality of inheritances into a single concrete
actuality.
The argument is therefore not that three vocabularies refer to a single
substance under different names; that would be a category error and a
poetic conflation.
The argument is that three vocabularies converge on a single \emph{role}
--- the role of sustaining integrated, unified being against the
decohering pressure of separation --- and identify that role as the
load-bearing structural feature of reality.
What the convergence detects is the role; the names belong to the
vocabularies, the role belongs to the world.

The table has a deliberate architecture.
Reading vertically, the four rows trace the same phenomenon across four
domains of inquiry --- from the formal physics of quantum coherence to
philosophical reflection, revealed scripture, and conscious experience.
Reading horizontally, each row names the same relational force in the
language native to its domain.
No row is more fundamental than the others; the Affinity Tensor is not
a more precise version of \mahabbat, any more than prehension is a more
precise version of love.
Each is the correct technical term for the relational force in the
vocabulary where that tradition operates.

The deepest implication of the semantic bridge is not terminological
but ontological.
If four domains, operating by different methods, all name the same force ---
if physicists call it entanglement and quantify it as the Affinity Tensor,
Whitehead calls it prehension or eros in the philosophical vocabulary of
process, the \bahai writings call it \mahabbat in the vocabulary of
revelation, and human beings call it love in the vocabulary of experience
--- then the most parsimonious explanation, by inference to the best
explanation, is that the force is real and singular, and that the
diversity of names reflects the diversity of the instruments trained
upon it, not a diversity of things.

This is not a new claim in the history of philosophy.
The intuition that love is a cosmic and not merely a personal force is
ancient (Empedocles' \emph{philia} and \emph{neikos} as cosmic principles;
Plato's \emph{Symposium}; Dante's ``love that moves the sun and other
stars'' at the close of the \emph{Paradiso}).
What this project adds is precision: a formal measure (the Affinity Tensor
$\alpha$), a testable prediction (the $\URA$ gate), and a three-tradition
independent confirmation that triangulates the target from different
directions simultaneously.
The argument is not that love is important.
The argument is that love --- understood as the force of relational affinity
that constitutes and sustains existence at every scale --- is what reality
is made of.
That argument is now formal enough to be tested, and it has been
independently arrived at by three traditions that developed independently
of one another.
