% ── SECTION 6b: THE RCF IN PHILOSOPHICAL LANGUAGE (~900 words) ───────────────

\section{The Relational Coherence Framework: Formalizing the Convergence}
\label{sec:rcf-phil}

The three preceding sections have made a philosophical argument.
Sections~\ref{sec:substance}--\ref{sec:whitehead} established that the
surviving interpretations of quantum mechanics are either explicitly
process ontologies or collapse on examination into relabeled versions
of one, and that Whitehead's framework provides the most fully
systematized articulation of the resulting picture.
Sections~\ref{sec:bahai}--\ref{sec:convergences} demonstrated that the \bahai
writings constitute an independent prior instantiation of the same structure
across six specific convergences.
Section~\ref{sec:semantic} showed that each tradition's technical vocabulary
names the same underlying phenomenon.

What converts this philosophical argument into a scientific contribution
with testable consequences is the \emph{Relational Coherence Framework} (\RCF):
a formal mathematical framework, developed in the companion physics
paper~\citep{Adams2026Physics}, that translates the central claim of
process ontology into the language of quantum mechanics and derives
quantitatively distinct predictions from it.
This section presents the \RCF\ in non-mathematical terms sufficient for
the philosophical argument.

\subsection{What the RCF Does}

The central claim of process ontology --- and of the three-tradition
convergence documented in this paper --- is that \emph{relational affinity
is the condition of existence}.
Things exist not as independent substances possessing intrinsic properties,
but as patterns of relational coherence: the ongoing relational bonds
that constitute a thing are what make it a thing.
When those bonds dissolve, the thing dissolves.

The \RCF\ translates this claim into formal quantum mechanics through
a single quantitative object: the \emph{Affinity Tensor} $\alpha(S,E)$.
For any quantum system $S$ in relation to its environment $E$, the
Affinity Tensor measures the degree to which $S$ retains relational
coherence with $E$ --- the magnitude of the quantum bonds that constitute
$S$'s participation in relational existence.
When $\alpha = 1$, $S$ is maximally coherent: fully participating in
the relational structure that constitutes quantum existence.
When $\alpha = 0$, $S$ is fully decohered: classically separate,
its relational bonds dissolved into the environment.
The range $[0,1]$ is not a metaphor.
It is a precisely defined, experimentally accessible, continuously
varying measure of what process philosophy calls the degree of a
thing's relational being.

\subsection{Four Propositions}

The \RCF\ comprises four propositions, each of which makes the
philosophical claim mathematically precise.

\paragraph{Property indefiniteness.}
A system with nonzero Affinity Tensor ($\alpha > 0$) does not possess
classical-style definite values for the full set of its complementary
observables, and (when entangled) does not possess definite local
properties at all.
This is not a statement about measurement uncertainty.
It is a statement about ontological structure: the system is still in
the process of becoming, still in the superposition of relational
possibilities that Whitehead would call its potential phase.
Classical-style definite local properties belong to decohered states;
the relational coherence preserved at $\alpha > 0$ and the classical
definiteness recovered at $\alpha = 0$ are, in this sense, mutually
exclusive --- exactly as process ontology predicts.

\paragraph{Relational determination.}
Definite properties emerge through relational interaction.
Under environmental coupling, the Affinity Tensor decays continuously
from its initial value toward zero.
As it decays, the system's relational potential is actualized into
definite outcomes through its interaction with the environment.
On our proposal, this is a mathematical formalization of what Whitehead
means by concrescence: the transition from potential to actuality through
relational process.

\paragraph{Coherence--affinity correspondence.}
The Affinity Tensor ranges continuously over $[0,1]$, with $\alpha = 1$
at maximum relational coherence and $\alpha = 0$ at full decoherence.
This means that relational affinity is not binary --- not simply present
or absent --- but a continuously graded quantity.
Existence, in the process ontological sense, admits of degrees.
The \bahai passage that defines nonexistence as the \emph{absence} of
affinity and attraction (Section~\ref{sec:bahai}) is, on this reading,
a statement about the $\alpha = 0$ limit: the point at which relational
coherence has been fully extinguished and the pattern of relational
events that constituted the thing has dissolved.

\paragraph{Conservation of relational potential.}
On the formal definition supplied in the companion
paper~\citep{Adams2026Physics}, the total relational potential of a
closed system--environment composite is conserved under unitary
evolution; what changes is its distribution.
Relational coherence that was locally concentrated in the system is
redistributed into correlations between the system and its environment,
in a manner consistent with the standard quantum-mechanical fact that
unitary evolution preserves global state purity while local von Neumann
entropies may grow.
Relational being, on this construal, is not created or destroyed but
transformed.
This is the formal expression of what the \bahai writings call the
creative foundation: the relational ground of existence is not consumed
by the process of becoming but redistributed through it, available for
the next occasion of relational actualization.

\subsection{The Relational Affinity Gate}

The \RCF\ has a direct application in quantum computing.
Standard treatments of two-qubit unitaries~\citep{NielsenChuang2000}
parameterize gates within the Weyl chamber of nonlocal
operations~\citep{Zhang2003WeylChamber}; the \RCF-derived gate occupies
a specific continuous trajectory through that chamber.
If relational coupling is a continuously variable resource rather than
a fixed quantity, then a two-qubit gate should expose a continuous
coupling parameter that can be tuned to the actual coherence budget of
a physical quantum processor.
The \emph{Relational Affinity Gate} ($\URA$), derived from the \RCF,
does exactly this: it is a continuously parameterized two-qubit unitary
that trades a fraction of its entangling capability for a proportionally
shorter operation time under noise.
On hardware where gate duration scales with coupling strength, the
companion paper predicts that $\URA$ at sub-maximal coupling will
achieve measurably higher process fidelity than fixed entangling gates
such as \textsc{cnot} or \textsc{iswap}~\citep{Adams2026Physics}.
At the per-gate level, the companion paper compares $\URA(\pi/8)$ ---
operating at half the duration of \textsc{iswap} --- against fixed
maximally entangling gates, reporting fidelity advantages of
$\Delta F \in [{+0.019},\,{+0.020}]$ at per-gate error rate
$p_{\rm base} = 0.05$ and $\Delta F \in [{+0.038},\,{+0.039}]$ at
$p_{\rm base} = 0.10$, consistent across depolarizing, dephasing, and
amplitude-damping noise models~\citep{Adams2026Physics}.
The principal technical contribution of the companion paper is the
\emph{Coherence-Budget Allocation Theorem}: given a target amount of
entanglement to be distributed across the layers of a quantum circuit
and a per-layer noise characterization, the theorem derives the unique
cost-minimizing exchange-angle schedule in closed form.
The theorem yields three falsifiable predictions: distribute
entanglement across layers rather than concentrate it in a single gate
(uniform-noise case); route heavier entanglement onto the quieter layers
(heterogeneous-noise case); and skip layers entirely when their local
noise rate exceeds a global threshold set by the coherence budget.
These predictions have been confirmed in multi-layer density-matrix
simulation, with the optimal schedule achieving fidelity advantages of
up to $\Delta F = +0.040$ over a concentrated full-entangler baseline
and $\Delta F = +0.025$ over a uniform-angle schedule on heterogeneous
noise~\citep{Adams2026Physics}.
This is not merely an engineering optimization.
It is a prediction derived from the process ontological claim that
relational affinity is a continuous, conservatively redistributed
resource --- a claim that the \RCF\ formalizes and that the $\URA$
gate tests experimentally.
The ``economical'' principle the theorem embodies --- extracting only
as much entanglement as the coherence budget can sustain, layer by
layer, routed toward the quietest substrates --- instantiates the
conservation of relational potential developed in the preceding
subsection and resonates with a broader emphasis on balance and measured
proportion that recurs across all three traditions: Whitehead's
redistribution of relational potential under unitary evolution, and the
\bahai teaching, articulated by \AbdulBaha, that integrated being is
sustained by attraction and agreement in proper measure rather than by
maximal force.

\subsection{The RCF as the Formal Expression of the Convergence}

The significance of the \RCF\ for the argument of this paper is this:
it demonstrates that the convergence documented in the preceding sections
is not merely philosophical.
The claim that existence is constituted by relational affinity --- made
by Whitehead through philosophical analysis, by the \bahai writings
through spiritual revelation, and by quantum mechanics through more than
six decades of experimental tests of Bell's inequalities --- is
sufficiently precise to generate a formal mathematical framework, a
novel quantitative object (the Affinity Tensor), a new two-qubit gate,
and testable predictions about gate fidelity under realistic noise.

The convergence is not a set of interesting similarities.
It is the identification of a real feature of reality --- the relational
constitution of existence --- that is precise enough to be formalized,
computed, and tested.
The \RCF\ is that formalization.
